---
title: "Data_Analysis"
output: html_document
date: "2025-04-15"
---
continual: age, calories, sugar, fat, BMI
ordinal: BMIcat, FoodSec, HHMembers
nominal: Gender, Race

under 0.05 is a signifciant number
```{r}
projectdatabase$BMIcat <- factor(projectdatabase$BMIcat)
```

```{r}
library(dplyr)
library(tidyr)
```
summary of calories, fat, and sugar intake:
```{r}
Diet <- c("Calories", "Fat", "Sugar")
summary(projectdatabase[Diet])
```

```{r}
#What are the means, standard deviations, and counts for total calories, total sugars, and total fat intake?
projectdatabase%>%
  drop_na(Calories)%>%
  summarize(mean(Calories), sd(Calories), count=n())
projectdatabase%>%
  drop_na(Sugar)%>%
  summarize(mean(Sugar), sd(Sugar), count=n())
projectdatabase%>%
  drop_na(Fat)%>%
  summarize(mean(Fat), sd(Fat), count=n())

```
```{r}
#Generate the mean, standard deviation, and counts for dietary protein and fat intake by BMIcat
projectdatabase%>%
  drop_na(Calories)%>%
  group_by(BMIcat)%>%
  summarise(mean(Calories), sd(Calories), count=n())
projectdatabase%>%
  drop_na(Fat)%>%
  group_by(BMIcat)%>%
  summarise(mean(Fat), sd(Fat), count=n())
projectdatabase%>%
  drop_na(Sugar)%>%
  group_by(BMIcat)%>%
  summarise(mean(Sugar), sd(Sugar), count=n())

```
descriptive statistics for BMI and Age
```{r}
projectdatabase%>%
  drop_na(BMI)%>%
  summarize(mean(BMI), sd(BMI), count=n())
projectdatabase%>%
  drop_na(Age)%>%
  summarize(mean(Age), sd(Age), count=n())
```
descriptive statistics for BMIcat, FoodSec, HHMembers, Race, and Gender
```{r}
library(questionr)
freq(projectdatabase$BMIcat, cum = TRUE)
freq(projectdatabase$FoodSec, cum = TRUE)
freq(projectdatabase$HHMembers, cum = TRUE)
freq(projectdatabase$Race, cum = TRUE)
freq(projectdatabase$Gender, cum = TRUE)
```
Bar charts showing categories by percentages for BMIcat, HHMembers, and FoodSec
```{r}
#Bar chart showing BMI categories by percentages:
BMIcat_table <- with(projectdatabase, table(BMIcat))
prop.table(BMIcat_table)
prop.table(BMIcat_table)*100
BMIcat_barchart <- as.data.frame(prop.table(BMIcat_table)*100)
ggplot(BMIcat_barchart, aes(BMIcat, Freq)) + geom_col()
```
```{r}
#Bar chart showing HHMembers by percentages:
HHMembers_table <- with(projectdatabase, table(HHMembers))
prop.table(HHMembers_table)
prop.table(HHMembers_table)*100
HHMembers_barchart <- as.data.frame(prop.table(HHMembers_table)*100)
ggplot(HHMembers_barchart, aes(HHMembers, Freq)) + geom_col()
```
```{r}
#Bar chart showing FoodSec by percentages:
FoodSec_table <- with(projectdatabase, table(FoodSec))
prop.table(FoodSec_table)
prop.table(FoodSec_table)*100
FoodSec_barchart <- as.data.frame(prop.table(FoodSec_table)*100)
ggplot(FoodSec_barchart, aes(FoodSec, Freq)) + geom_col()
```
```{r}
#Bar chart showing Race by percentages:
Race_table <- with(projectdatabase, table(Race))
prop.table(Race_table)
prop.table(Race_table)*100
Race_barchart <- as.data.frame(prop.table(Race_table)*100)
ggplot(Race_barchart, aes(Race, Freq)) + geom_col()
```
```{r}
#Bar chart showing Gender by percentages:
Gender_table <- with(projectdatabase, table(Gender))
prop.table(Gender_table)
prop.table(Gender_table)*100
Gender_barchart <- as.data.frame(prop.table(Gender_table)*100)
ggplot(Gender_barchart, aes(Gender, Freq)) + geom_col()

BMI_gender <- projectdatabase%>%
  drop_na(BMI)%>%
  group_by(Gender)%>%
  summarize(n=n(), mean=mean(BMI), sd=sd(BMI))
BMI_gender

ggplot(BMI_gender, aes(x=reorder(Gender, mean), y=mean)) + geom_bar(stat = "identity", fill="lightblue") + geom_errorbar(aes(ymin=mean-sd, ymax=mean+sd), width=0.2)
labs(x="Gender",
     y="",
     title = "Mean BMI",
     subtitle = "with standard deviation")
```
Historgram of BMI:
```{r}
#a. simple histogram: ggplot(data, aes(x=contvar)) + geom_histogram() + labs(title="", x="", y="")
#construct a histogram for BMI:
ggplot(projectdatabase, aes(x=BMI)) + 
  geom_histogram() + labs(title = "histogram of BMI")
```
Histogram of Age:
```{r}
#a. simple histogram: ggplot(data, aes(x=contvar)) + geom_histogram() + labs(title="", x="", y="")
#construct a histogram for Age:
ggplot(projectdatabase, aes(x=Age)) + 
  geom_histogram() + labs(title = "histogram of Age")
```
Histogram of Sugar:
```{r}
ggplot(projectdatabase, aes(x=Sugar)) + 
  geom_histogram() + labs(title = "histogram of Sugar")
```
Histogram of Calories:
```{r}
ggplot(projectdatabase, aes(x=Calories)) + 
  geom_histogram() + labs(title = "histogram of Calories")
```
Histogram of Fat:
```{r}
ggplot(projectdatabase, aes(x=Fat)) + 
  geom_histogram() + labs(title = "histogram of Fat")
```

```{r}
#mytable <- with(data, table(variable))
race <- with(projectdatabase, table(Race))
race
#turn the above table into propotions with prop.table(mytable)
prop.table(race)
#turn the table into percentages using prop.table(mytable)*100
prop.table(race)*100

gender <- with(projectdatabase, table(Gender))
gender
prop.table(gender)
prop.table(gender)*100
```
```{r}
projectdatabase%>%
  group_by(Gender)%>%
  drop_na(BMI)%>%
  summarise(mean(BMI), sd(BMI), count=n())
```
```{r}
table(projectdatabase$Race, projectdatabase$BMIcat)

ggplot(projectdatabase, aes(x=BMIcat, fill=Race)) + geom_bar(position = "stack") + labs(title = "stacked bar chart", x="BMI Categories by Race", y="counts")

ggplot(projectdatabase, aes(x=BMIcat, fill=Race)) + geom_bar(position = "stack") + facet_wrap(~Race) + labs(title = "stacked bar chart", x="BMI Categories by Race", y="counts") + theme(axis.text.x = element_text(angle=45, hjust=1, size=10))

ggplot(projectdatabase, aes(x=BMIcat, fill=Race)) + geom_bar(position = "stack") + facet_grid(~Race) + labs(title = "stacked bar chart", x="BMI Categories by Race", y="counts") + theme(axis.text.x = element_text(angle=45, hjust=1, size=10))

ggplot(projectdatabase, aes(x=BMIcat, fill=Race)) + geom_bar(position = "dodge") + labs(title="grouped bar chart", x="BMI categories by race", y="counts")

```
```{r}
#install.packages("questionr")
library(questionr)
#use the following code to generate a frequency table for each ordinal variable.
#freq(data$variable, cum = TRUE, na.omit(data$variable)) (do not use na.omit for BMIcat)
freq(projectdatabase$Age, cum = TRUE, na.omit(projectdatabase$Age))
```
```{r}
#Calories frequency table
freq(projectdatabase$Calories, cum = TRUE, na.omit(projectdatabase$Calories))
```
```{r}
#Sugar frequency table
freq(projectdatabase$Sugar, cum = TRUE, na.omit(projectdatabase$Sugar))
```
```{r}
#Fat frequency table
freq(projectdatabase$Fat, cum = TRUE, na.omit(projectdatabase$Fat))
```
```{r}
#BMI cat frequency table
freq(projectdatabase$BMIcat, cum = TRUE)
```
```{r}
ggplot(projectdatabase, aes(x=BMIcat, fill=Gender)) + geom_bar(position = "dodge") + labs(title="grouped bar chart", x="BMI categories by gender", y="counts")
```
Correlation test: 
```{r}
projectdatabase %>%
  ggplot(., aes(x=BMI, y=Sugar)) + geom_point(color="violet") + geom_smooth(method=lm, color="purple") + labs(x="BMI", y="Daily Sugar Intake") + theme_bw()
#correlation test code: cor.test(data$DepVar, data$IndVar, method="pearson")
cor.test(projectdatabase$BMI, projectdatabase$Sugar, method="pearson")
```
Linear regression model 1: scatter plot
```{r}
#model_name <- lm(DepVar~IndVar, data=)
#summary(model_name)

BMI_Sugar <- lm(BMI~Sugar, projectdatabase)
summary(BMI_Sugar)
```
```{r}
#scatter plot code: #ggplot(aes(x=DepVar, y=IndVar), data=) + geom_point(color="") + geom_smooth(method=lm, color="") + labs(x="", y="") + theme_bw()
ggplot(aes(x=BMI, y=Sugar), data=projectdatabase) + geom_point(color="magenta") + geom_smooth(method=lm, color="blueviolet") + labs(x="BMI", y="Sugar") + theme_bw()
#correlation test code: cor.test(data$DepVar, data$IndVar, method="pearson")
cor.test(projectdatabase$BMI, projectdatabase$Sugar, method="pearson")
```

Linear regression model 2: 
```{r}
#model_name <- lm(DepVar~IndVar, data=)
#summary(model_name)

BMI_Calories <- lm(BMI~Calories, projectdatabase)
summary(BMI_Calories)
```
```{r}
#scatter plot code: #ggplot(aes(x=DepVar, y=IndVar), data=) + geom_point(color="") + geom_smooth(method=lm, color="") + labs(x="", y="") + theme_bw()
ggplot(aes(x=BMI, y=Calories), data=projectdatabase) + geom_point(color="lightgreen") + geom_smooth(method=lm, color="darkgreen") + labs(x="BMI", y="Calories") + theme_bw()
#correlation test code: cor.test(data$DepVar, data$IndVar, method="pearson")
cor.test(projectdatabase$BMI, projectdatabase$Calories, method="pearson")
```
Linear regression model 3: 
```{r}
#model_name <- lm(DepVar~IndVar, data=)
#summary(model_name)

BMI_Fat <- lm(BMI~Fat, projectdatabase)
summary(BMI_Fat)
```
```{r}
#scatter plot code: #ggplot(aes(x=DepVar, y=IndVar), data=) + geom_point(color="") + geom_smooth(method=lm, color="") + labs(x="", y="") + theme_bw()
ggplot(aes(x=BMI, y=Fat), data=projectdatabase) + geom_point(color="dodgerblue3") + geom_smooth(method=lm, color="navy") + labs(x="BMI", y="Fat") + theme_bw()
#correlation test code: cor.test(data$DepVar, data$IndVar, method="pearson")
cor.test(projectdatabase$BMI, projectdatabase$Fat, method="pearson")
```

Correlation test:
```{r}
projectdatabase %>%
  ggplot(., aes(x=BMI, y=Calories)) + geom_point(color="lightgreen") + geom_smooth(method=lm, color="darkgreen") + labs(x="BMI", y="Daily Caloric Intake") + theme_bw()
#correlation test code: cor.test(data$DepVar, data$IndVar, method="pearson")
cor.test(projectdatabase$BMI, projectdatabase$Calories, method="pearson")
```

```{r}
BMI_Calories <- lm(BMI~Calories, projectdatabase)
summary(BMI_Calories)
```
Correlation test:
```{r}
projectdatabase %>%
  ggplot(., aes(x=BMI, y=Fat)) + geom_point(color="lightblue") + geom_smooth(method=lm, color="navyblue") + labs(x="BMI", y="Daily Fat Intake") + theme_bw()
#correlation test code: cor.test(data$DepVar, data$IndVar, method="pearson")
cor.test(projectdatabase$BMI, projectdatabase$Fat, method="pearson")
```

```{r}
BMI_Fat <- lm(BMI~Fat, projectdatabase)
summary(BMI_Fat)
```

Chi-Square test:bar graph
```{r}
#a. Generate a two-way frequency table of BMIcat and HHMembers:
#table_name <- xtabs(~ CatVar1 + CatVar2, data=)
#print the table
#show the table in percentages: prop.table(table_name)*100
BMIcat_HHMembers_table <- xtabs(~BMIcat + HHMembers, data=projectdatabase)
BMIcat_HHMembers_table
prop.table(BMIcat_HHMembers_table)*100

chisq.test(BMIcat_HHMembers_table)

```
bar graph for Chi-square test
```{r}
ggplot(projectdatabase, aes(x=BMIcat, fill=HHMembers)) + geom_bar(position = "dodge") + labs(title="grouped bar chart", x="BMI categories by HHMembers", y="counts")
```

T-Test: bar graph, box plot
```{r}
#2 descriptive statistics of sedentary grouped by health insurance
projectdatabase%>%
  drop_na(Calories)%>%
  group_by(Gender)%>%
  summarize(mean(Calories), sd(Calories), count=n())
#3 test to see if the variance of the continuous variable are equal across different levels of the categorical variable: var.test(contvar~catvar, data=, )
var.test(Calories~Gender, data=projectdatabase, alternative="two.sided")
#4 t-test: t.test(contvar~catvar, var.equal=FALSE, data=, na.rm=TRUE)
t.test(Calories~Gender, var.equal=FALSE, data=projectdatabase, na.rm=TRUE)
```
the above T-test does not indicate a significant P-value. 

graph for t-test 1:(gender and Calories)
```{r}
ggplot(projectdatabase, aes(x=Gender, fill=Calories)) + geom_bar(position = "dodge") + labs(title="grouped bar chart", x="Gender by Calories", y="counts")+ geom_bar(color = "lightgreen", fill="darkgreen")
```
T-test again:
```{r}
#2 descriptive statistics of sedentary grouped by health insurance
projectdatabase%>%
  drop_na(Sugar)%>%
  group_by(Gender)%>%
  summarize(mean(Sugar), sd(Sugar), count=n())
#3 test to see if the variance of the continuous variable are equal across different levels of the categorical variable: var.test(contvar~catvar, data=, )
var.test(Sugar~Gender, data=projectdatabase, alternative="two.sided")
#4 t-test: t.test(contvar~catvar, var.equal=FALSE, data=, na.rm=TRUE)
t.test(Sugar~Gender, var.equal=FALSE, data=projectdatabase, na.rm=TRUE)
```
the above T-test does not indicate a significant P-value. 

graph for the second t-test: (gender and sugar)
```{r}
ggplot(projectdatabase, aes(x=Gender, fill=Sugar)) + geom_bar(position = "dodge") + labs(title="grouped bar chart", x="Gender by Sugar", y="counts")+ geom_bar(color = "purple", fill="violet")
```
t-test again:
```{r}
#2 descriptive statistics of sedentary grouped by health insurance
projectdatabase%>%
  drop_na(Fat)%>%
  group_by(Gender)%>%
  summarize(mean(Fat), sd(Fat), count=n())
#3 test to see if the variance of the continuous variable are equal across different levels of the categorical variable: var.test(contvar~catvar, data=, )
var.test(Fat~Gender, data=projectdatabase, alternative="two.sided")
#4 t-test: t.test(contvar~catvar, var.equal=FALSE, data=, na.rm=TRUE)
t.test(Fat~Gender, var.equal=FALSE, data=projectdatabase, na.rm=TRUE)
```
the above T-test does not indicate a significant P-value. 

graph for the third t-test: (gender and fat)
```{r}
ggplot(projectdatabase, aes(x=Gender, fill=Fat)) + geom_bar(position = "dodge") + labs(title="grouped bar chart", x="Gender by Fat", y="counts") + geom_bar(color = "blue", fill="navy")


```

ANOVA test:
```{r}
#1. find summary statistics of sample sizes, means, standard deviations, and 95% confidence intervals
#age is the dependent variable and BMIcat is the independent variable. the higher the F falue the more significant ANOVA is. 
plotdata<- projectdatabase%>%
  drop_na(Calories)%>%
  group_by(BMIcat)%>%
  summarize (n=n(),
             mean = mean(Calories),
             sd=sd(Calories),
             ci=qt(0.975, df=n-1)*sd/sqrt(n))
plotdata
#2. test for group differences (ANOVA)
fit <- aov(Calories~BMIcat, data=projectdatabase)
summary(fit)
#3. Plot group means and confidence intervals to visualize these group differences.
library(ggplot2)
ggplot(plotdata, aes(x=BMIcat, y=mean, group = 1)) +
  geom_point(size=2, color="darkgreen") +
             geom_errorbar(aes(ymin=mean-ci,
                               ymax=mean+ci),
                           width=.1) +
               theme_bw() +
               labs(x="BMI category",
                    y="Calories",
                    title = "BMI category and Caloric intake with 95% CI")
```
ANOVA again:
```{r}
#1. find summary statistics of sample sizes, means, standard deviations, and 95% confidence intervals
#age is the dependent variable and BMIcat is the independent variable. the higher the F falue the more significant ANOVA is. 
plotdata2<- projectdatabase%>%
  drop_na(Fat)%>%
  group_by(BMIcat)%>%
  summarize (n=n(),
             mean = mean(Fat),
             sd=sd(Fat),
             ci=qt(0.975, df=n-1)*sd/sqrt(n))
plotdata2
#2. test for group differences (ANOVA)
fit <- aov(Fat~BMIcat, data=projectdatabase)
summary(fit)
#3. Plot group means and confidence intervals to visualize these group differences.
library(ggplot2)
ggplot(plotdata2, aes(x=BMIcat, y=mean, group = 1)) +
  geom_point(size=2, color="navyblue") +
             geom_errorbar(aes(ymin=mean-ci,
                               ymax=mean+ci),
                           width=.1) +
               theme_bw() +
               labs(x="BMI category",
                    y="Fat",
                    title = "BMI category and Fat intake with 95% CI")
```
ANOVA again:
```{r}
#1. find summary statistics of sample sizes, means, standard deviations, and 95% confidence intervals
#age is the dependent variable and BMIcat is the independent variable. the higher the F falue the more significant ANOVA is. 
plotdata3<- projectdatabase%>%
  drop_na(Sugar)%>%
  group_by(BMIcat)%>%
  summarize (n=n(),
             mean = mean(Sugar),
             sd=sd(Sugar),
             ci=qt(0.975, df=n-1)*sd/sqrt(n))
plotdata3
#2. test for group differences (ANOVA)
fit <- aov(Sugar~BMIcat, data=projectdatabase)
summary(fit)
#3. Plot group means and confidence intervals to visualize these group differences.
library(ggplot2)
ggplot(plotdata3, aes(x=BMIcat, y=mean, group = 1)) +
  geom_point(size=2, color="violet") +
             geom_errorbar(aes(ymin=mean-ci,
                               ymax=mean+ci),
                           width=.1) +
               theme_bw() +
               labs(x="BMI category",
                    y="Sugar",
                    title = "BMI category and Sugar intake with 95% CI")
```

odds ratio:
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
install.packages("epitools")
library(epitools)
install.packages("ggmosaic")
library(ggmosaic)
```
odds ratio:
```{r}
BMIcat_Gender_odds_table <- table(projectdatabase$BMIcat, projectdatabase$Gender)

oddsratio(BMIcat_Gender_odds_table)

ggplot(projectdatabase) + geom_mosaic(aes(x=product(BMIcat), fill=Gender), na.rm=TRUE, show.legend=FALSE) + labs(title="Gender by BMIcat", x="BMIcat", y="Gender")
```
the above odds ratio is not significant. 